Solving an Exam Scheduling Problem Using a Genetic Algorithm

### 2009-12-02 Paper/paper.tex
Paper update

\documentclass[letterpaper]{article}
\usepackage{aaai}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage{url}
\usepackage[table]{xcolor}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{float}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PDFMARK for TeX and GhostScript
% Uncomment and complete the following for metadata if
% your paper is typeset using TeX and GhostScript (e.g
% if you use .ps or .eps files in your paper):
% \special{! /pdfmark where
% {pop} {userdict /pdfmark /cleartomark load put} ifelse
% [ /Author (John Doe, Jane Doe)
% /Title (Paper Title)
% /Keywords (AAAI, artificial intelligence)
% /DOCINFO pdfmark}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PDFINFO for PDFTeX
% Uncomment and complete the following for metadata if
% your paper is typeset using PDFTeX
\pdfinfo{
  /Title (Solving an Exam Scheduling Problem Using a Genetic Algorithm)
  /Subject (Input The Proceedings Title Here)
  /Author (Dave, Kordalewski;
           Caigu, Liu;
           Kevin, Salvesen;)
} 
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Uncomment only if you need to use section numbers
% and change the 0 to a 1 or 2
% \setcounter{secnumdepth}{0}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

  The total number of different possible schedules in any problem instance is
  $|C|^{|R|dt}$ which is typically far too large for brute force search.
  For example, one instance that we examine (scheduling the fall semester
  exams at the University of Toronto in 2009) involves 603 courses, 7 days,
  8 times, 43 rooms, and 21945 students. This instance admits approximately 
  $10^{6695}$ different possible schedules.

@@

\section{Approach and Implementation}
  \subsection{Differences From Natural Exam Scheduling Problems}
    The problem we have constructed here is a compromise between the most
    mathematically pure problem with the same sort of properties (which might ignore the
    rooms and related constraints completely) and the problem that must be solved in the
    real world when scheduling thousands of students, which has some important differences
    from this problem. For example, we don't deal with constraints such as room capacity
    or exams with varying length.

  \subsection{Tools used}
    Our application was entirely written in Java. We also wrote another Java application 
    to generate problem instance files for testing our algorithms. We used Mathematica 
    to plot all of our graphs.  Additionally, we use Google code's source code
    repository\footnote{\url{http://code.google.com/p/csc384-genetic-algorithms-project/}} 
    and Google docs to collaborate on this report.

  \subsection{Fitness Function}
    In order to assign each schedule a fitness, we need to define the functions 
    $q_{student}$ and $q_{room}$. They need to capture the idea of determining how
    much each student or room likes their given timetable. We do this by calculating 
    the penalty incurred by a room, which is a measure of it having properties that 
    are unpleasant to the student or room, and then using the formula

@@

    This function will be in the desired range or [0,1].

    For students, we accrue a penalty of 50 for every time that is repeated in a 
    timetable, 5 when there are exams in 2 consecutive times, 3 when 2 exams happen 
    on the same day, and 0.5 when the student has exams on consecutive days. This 
    accords reasonably well with real students' hopes for their exam timetables.

    Rooms have the same penalty of 50 when it is used by two exams at the same time, 
    5 if it hosts exams in consecutive time slots (reasoning that the invigilators 
    need time to let the students out and in and prepare for the next exam) and a 
    penalty of 0.5 when a room is empty for an entire day between uses (reasoning 
    that the room requires some amount of modifications to be used for exams, and 
    it is difficult to use it for other purposes if it will still be used for exams 
    soon.)

@@

    \subsubsection{General Description}
      The genetic algorithm maintains a population of schedules and constructs subsequent
      generations by adding new random schedules, making copies of high quality schedules 
      in the present generation, applies the mutation operator to schedules, or uses the
      crossover operator on 2 schedules. When performing mutation or crossover, the 
      schedules are selected with a probability proportional to their fitness, allowing 
      the schedules with high quality to be selected more frequently and make more 
      ``offspring''.

@@

@@

      The second set was generated with a sort of correlation between the courses that each 
      student took, rendering the dataset more true to the sort of data that would be found 
      in real-world situations. The set of courses that each student takes was selected such 
      that we could expect that 70\% of their courses were courses in their major. This 
      simulates the way in which we expect student bodies to choose their courses, though 
      the true data is likely to exhibit far more structure, due to program requirements, 
      notorious professors, etc.

      The 3 instances' size vary from manageable to large. The smallest could potentially be 
      solved exactly, with methods not under consideration here, while the others have 
      search spaces much too large for any hope of finding a perfect solution. For the 
      largest, which simulates a UofT exam schedule, a single evaluation of the fitness 
      function takes 0.2 seconds on on a typical desktop computer. This makes it difficult 
      to thoroughly investigate this instance.

@@

	  The large data set was created with parameters inspired from the University of 
	  Toronto's Faculty of Arts and Science.

@@

      \paragraph{Generation Parameters}
        We undertook a search of the space of parameters for forming new generations.

        Our results can be seen on figure 1, where each data point of the graph 
        represents the average of 5 trials. To make the scale clearer, the colour 
        on the surface corresponds to the fitness, where the highest fitnesses 
        are red. The number of copies is fixed at 1, and the proportions of 
        mutations and crossovers are shown on the horizontal axes.

@@

      \paragraph{Size Parameters}
        We also performed tests to gauge the optimal population size and number of generations 
        to calculate. We determined that very small values for either the population or number 
        of generations result in very poor generated solutions. A population of around 50 is 
        sufficient for good diversity to be retained in the population, and having a number of 
        generations relatively large compared to the population size seemed to be effective.

        We gathered data from the small and medium problem instances, but the precise qualities 
        provided by particular parameters will not necessarily hold in larger instances. The 
        optimal population size and number of generations (for a fixed number of fitness 
        evaluations) is likely to vary for larger problem instances, but it seemed clear that 
        the number of generations should be set somewhat higher than the population.

@@

\section{Conclusion}
  Any monotone function can be applied to the fitness values to obtain identical 
  results for the 3 non-GA methods, which only need to choose the better fitness 
  and do not use the fitness values in any important way. The GA, however, uses 
  the fitness values to decide the probability of schedules being selected and 
  modified for subsequent generations. It might be possible to reduce or increase 
  the probability of poor functions being selected for mutation or crossover.

>>>>

\documentclass[letterpaper]{article}
\usepackage{aaai}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage{url}
\usepackage[table]{xcolor}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{float}
\usepackage{natbib}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PDFMARK for TeX and GhostScript
% Uncomment and complete the following for metadata if
% your paper is typeset using TeX and GhostScript (e.g
% if you use .ps or .eps files in your paper):
% \special{! /pdfmark where
% {pop} {userdict /pdfmark /cleartomark load put} ifelse
% [ /Author (John Doe, Jane Doe)
% /Title (Paper Title)
% /Keywords (AAAI, artificial intelligence)
% /DOCINFO pdfmark}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PDFINFO for PDFTeX
% Uncomment and complete the following for metadata if
% your paper is typeset using PDFTeX
\pdfinfo{
  /Title (Solving an Exam Scheduling Problem Using a Genetic Algorithm)
  /Subject (Input The Proceedings Title Here)
  /Author (Dave, Kordalewski;
           Caigu, Liu;
           Kevin, Salvesen;)
} 
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Uncomment only if you need to use section numbers
% and change the 0 to a 1 or 2
% \setcounter{secnumdepth}{0}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

\section{Approach and Implementation}
  \subsection{Differences From Natural Exam Scheduling Problems}
    The problem we have constructed here is a compromise between more mathematically
    pure problems with the same sort of properties (which might ignore the rooms' 
    constraints completely) and the problem that must be solved in the real world 
    when scheduling thousands of students, which has some important differences from 
    this problem. For example, we don't deal with constraints like room capacity or 
    exams with varying length.

  \subsection{Tools used}
    Our application was entirely written in Java. We also wrote another Java application 
    to generate problem instance files for testing our algorithms. We used Mathematica 
    to plot all of our graphs.  Additionally, we use Google code's subversion
    repository\footnote{\url{http://code.google.com/p/csc384-genetic-algorithms-project/}},
    Google docs and Google Wave to collaborate on this report.

  \subsection{Fitness Function}
    In order to assign each schedule a fitness, we need to define the functions
    $q_{student}$ and $q_{room}$. They need to capture the idea of determining 
    how much each student or room likes their given timetable. We do this by giving 
    penalties to schedules when they have properties that are disliked by the 
    student or room, and then using the formula

@@

    This function will be in the desired range of [0,1].

    We assign 50 point penalties for inconsistent schedules, 3-5 points when 
    students are assigned multiple exams on the same day, and 0.5 when students 
    don't have a rest day between exams. Rooms have the same stiff 50 point penalty 
    for inconsistencies, 5 for hosting exams in consecutive time slots, and 0.5 for 
    any days without exams between the days it is first used and last used.

@@

    \subsubsection{General Description}
      The genetic algorithm maintains a population of schedules and constructs subsequent 
      generations by adding new random schedules, making copies of high quality schedules 
      in the present generation, applies the mutation operator to schedules, or uses the 
      crossover operator on 2 schedules. When performing mutation or crossover, the 
      schedules are selected with a probability proportional to their fitness, allowing 
      the schedules with high quality to be selected more frequently and make more 
      ``offspring".

@@

    \subsubsection{Parallel Computations}
      Of the 4 methods, only mutate search is not trivially parallelizable. A plausible 
      parallel version of mutate search would generate as many mutated schedules at 
      every step as processors are available and select the fittest result for the next 
      set of mutations. We expect that this would increase the rate at which the best 
      found fitness increases.

@@

    The second set was generated with a correlation between the courses that each 
    student took, rendering the data-set more true to the sort of data that would 
    be found in real-world situations. The set of courses that each student takes 
    was selected so that around 70\% of their courses were in their (randomly selected) 
    major. This simulates the way in which we expect student bodies to choose their 
    courses, though the true data is likely to exhibit far more structure, due to 
    program requirements, notorious professors, etc.

    The sized of our 3 instances' vary from manageable to large. The total number 
    of different possible schedules in any problem instance is $|C|^{|R|dt}$ which 
    is far too large to conduct a brute force search. The smallest could potentially 
    be solved exactly, with methods not under consideration here, while the others 
    have search spaces much too large for any hope of finding a perfect solution. 
    For the largest, which simulates a UofT exam schedule, a single evaluation of 
    the fitness function takes 0.2 seconds on on a typical desktop computer. This 
    makes it difficult to thoroughly investigate this instance.

@@

	The large data set was created with parameters corresponding to the fall 2009 
	University of Toronto exam schedule, which is the size we expect for a large 
	real-world problem instance.

@@

      \paragraph{Generation Parameters}
        We undertook a search of the space of parameters for forming new generations. 
        Our results for different combinations of parameters can be seen in figure 1. 
        Each point has been averaged over 5 trials. The colour on the plotted surface 
        corresponds to the fitness, where the highest fitnesses are red. The number of 
        copies is fixed at 1, and the proportion of mutations and crossovers are shown 
        on the horizontal axes.

@@

@@

\section{Conclusion}
  Any monotone function can be applied to the fitness values to obtain identical 
  results for the 3 non-GA methods, which only need to choose the better fitness 
  and do not use the fitness values in any important way. The GA, however, uses 
  the fitness values to decide the probability of schedules being selected and 
  modified for subsequent generations. It might be possible to reduce or increase 
  the probability of poor functions being selected for mutation or crossover.
  \citeauthor*{pdf:solvingexam} \citeyearpar{pdf:solvingexam} for example, got 
  favourable results by squaring the fitness.
